\section{Architecture and Extensibility}
\label{sec:architecture}

Based on the astrolabe store defined in \S\ref{sec:data-model}, we developed a Tauri desktop application for reading, visualizing, and interacting with the data structure. We use a Python backend (FastAPI) to implement the core data model and a React frontend (Next.js) for visualization and authoring; they communicate via REST. All data flows through the astrolabe store as the single source of truth.

\subsection{Core Layer}

The core is deliberately minimal. It implements entry CRUD, hash computation, and automatic hash propagation (\S\ref{sec:data-model}), exposed via a REST API. The \texttt{record} field is treated as an opaque string: the core knows nothing about ``definitions,'' ``theorems,'' or ``proofs.'' All domain-specific logic is delegated to plugins.

The frontend core provides three workspace panels (Read View, Network View, Detail View), an AI chat assistant via Tauri IPC, and a preprocessor that converts \LaTeX-style entry macros (\verb|\entryblock|, \verb|\entryref|) into interactive components (\S\ref{sec:visualization}).

\subsection{Plugin Architecture}

Each plugin implements the \texttt{AstrolabePlugin} interface and is registered in a runtime-toggleable store. A plugin can contribute backend API endpoints (under \texttt{/api/plugins/}), frontend UI panels, and record conventions, all without modifying the core. Plugins are typed into three categories:
\begin{itemize}
  \item \textbf{Source}: imports external data into the store. Examples: \LaTeX\ parser, BibTeX importer, Lean~4 artifact parser.
  \item \textbf{Endo}: transforms the store in place. Examples: network metric computation, LLM embedding, metadata enrichment.
  \item \textbf{Sink}: reads the store and produces output. Examples: visualization, export, statistical summary.
\end{itemize}

A typical workflow composes plugins as Source $\to$ Endo $\to \cdots \to$ Endo $\to$ Sink, with the astrolabe store as the single intermediate representation at every stage. No format conversion is needed between plugins.

\subsection{Desktop Shell}

Tauri wraps the web frontend into a native desktop application using the system webview, resulting in small binaries and low memory usage. The Python backend is bundled as a sidecar process managed by the Tauri runtime.
