\section{Background and Related Work}
\label{sec:background}


This section surveys four lines of work that Astrolabe draws from and positions against: formalization planning tools (blueprints), network science applied to mathematical knowledge, content-addressing for formal mathematics, categorical approaches to data modeling, and network-oriented personal knowledge management tools.


\subsection{Blueprints as Formalization Planning Tools}
\label{sec:bg-blueprints}

Scholze's Liquid Tensor Experiment catalyzed the idea that large formalization projects benefit from explicit dependency planning. Massot's leanblueprint~\cite{Massot2020} provided the first systematic tool for this: a \LaTeX{} document declares theorems and their dependencies via \texttt{\textbackslash uses}, and a web frontend renders a colored dependency graph tracking formalization progress. The Polynomial Freiman--Ruzsa project~\cite{Tao2023} demonstrated the approach at scale, with blueprint-first planning enabling distributed collaboration among dozens of contributors.

More recently, LeanArchitect (Zhu, 2026) took a complementary approach: instead of manually writing blueprint metadata in \LaTeX{}, it extracts dependency information directly from compiled Lean code, eliminating the dual-maintenance burden of keeping \LaTeX{} and Lean in sync. The two tools address different phases of a project---leanblueprint for upfront planning, LeanArchitect for documentation of existing code.

Both tools share a structural limitation: \texttt{\textbackslash uses} is the only relation type. There is no distinction between statement-level dependencies and proof-technique dependencies, no mechanism for network analysis (centrality, community structure), and no way for formalization artifacts to flow back into the planning layer. These are not flaws in leanblueprint or LeanArchitect---they are architectural absences that motivate a different kind of tool.


\subsection{Network Science Perspectives on Mathematical Knowledge}
\label{sec:bg-network-science}

In prior work~\cite{LiPengSeverini2026}, we analyzed the dependency structure of Lean~4's Mathlib library as a multi-layer network, extracting graphs at the declaration, module, and namespace levels. The analysis revealed structural features invisible at the individual-declaration level: 74.2\% of declaration-level edges are compiler-derived (not human-authored), cross-namespace dependencies account for 50.9\% of all edges, and the library exhibits hierarchical layering with 153, 84, and 8 layers at the three respective levels. These are objective structural properties of the codebase that can inform engineering decisions---the \texttt{shake} linter, typeclass caching, and parallel compilation all operate on exactly this kind of structural signal.

Network science has a long history of extracting actionable signals from knowledge structures. Citation network analysis, dating to Garfield~\cite{Garfield1955} and de Solla Price~\cite{Price1965}, demonstrated that the topology of scholarly references encodes influence patterns invisible in individual papers. Knowledge graph completion methods~\cite{Bordes2013} predict missing relations from graph structure. Concept map analysis in education~\cite{Novak2008} uses graph properties to assess the quality of student understanding. The common theme: graph structure carries information that node-level analysis misses.

The key insight for our work is that these network signals can guide formalization strategy. A node with high betweenness centrality in a formalization dependency graph is a bottleneck: formalizing it unlocks the most downstream work. Community detection identifies clusters of interdependent results that can be formalized as a unit. But these signals are only meaningful when edges carry type information---computing betweenness on a graph where ``statement dependency'' and ``proof-technique dependency'' are conflated into a single edge type produces noisy, uninterpretable rankings.

No existing tool computes network analysis metrics on formalization dependency graphs. doc-gen4 and LeanDepViz render flat, single-type dependency visualizations. leanblueprint's graph is a progress tracker, not an analytical tool. Astrolabe's analysis plugin system addresses this gap: network algorithms run on the typed obj/mor graph, producing centrality, community, and bottleneck scores that map to visual properties (node size, color grouping) in the force-directed Canvas.


\subsection{Content-Addressing and Decentralized Formalization}
\label{sec:bg-content-addressing}

The Stacks Project~\cite{deJong} pioneered stable identifiers for mathematical results: each lemma receives a tag that persists across revisions, enabling durable cross-references. This is content-addressing at the item level---but the data model is flat (tags identify items, not relationships), and there is no mechanism for computing tags from mathematical content. The IPFS/CID paradigm in software engineering demonstrates that content-addressed identifiers enable decentralization: Nix derivations, IPFS objects, and Git commits all derive identity from content, enabling trustless verification and deduplication.

In prior work~\cite{LiParisSeverini2026}, we developed a content-addressed hashing pipeline for Lean declarations: canonical reduction (stripping metadata, normalizing universe variables) followed by expression serialization and SHA-256 hashing. This enables cross-library deduplication---two independently developed libraries that prove the same theorem produce identical hashes. Astrolabe's identifier design draws from this work but separates two layers: the core layer uses random 12-character hex IDs for graph-internal stability (natural-language content has no canonical form), while the formalization plugin layer can attach content-addressed hashes computed from type signatures, enabling cross-instance merging when formal content is available.


\subsection{Category Theory in Data Modeling}
\label{sec:bg-category-theory}

Spivak and Kent's \emph{Ologs}~\cite{Spivak2012} are the most direct theoretical precursor to our work. An olog is built on a core triad: \emph{types} (objects representing classes of things), \emph{aspects} (morphisms representing functional relations between types), and \emph{facts} (commutative diagrams asserting that two paths through the olog yield the same result). This triad gives ologs three capabilities that plain directed graphs lack: commutative diagrams express path equivalence---a semantic notion beyond simple connectivity; functors provide precise mappings between knowledge bases; and ologs convert directly to relational database schemas.

Particularly relevant to our work is Spivak's treatment of \emph{communication between ologs} (\cite{Spivak2012}, Section~4). Two ologs $\mathcal{S}_1$ and $\mathcal{S}_2$ maintained by different authors can be connected via a common-ground span $\mathcal{S}_1 \xleftarrow{H_1} \mathcal{S} \xrightarrow{H_2} \mathcal{S}_2$, where the functors $H_1, H_2$ map shared concepts to their respective representations. This structure enables \emph{knowledge fusion}---combining information from independent sources while preserving structural consistency. Our vision of merging multiple Astrolabe instances via content-addressed hashes (Section~\ref{sec:dm-identifiers}) is an engineering realization of this categorical construction: the hash serves as the common ground, and the import functors serve as the connecting maps.

Spivak also develops a subtype mechanism via \emph{injective aspects} and pullbacks (\cite{Spivak2012}, Sections~5.4--5.5): a hook arrow $A \hookrightarrow B$ denotes that $A$ is a subtype of $B$, and pullbacks compute intersections of subtypes (e.g., ``wealthy customers'' $\cap$ ``loyal customers'' = pullback over ``customers''). Our \texttt{inclusion} morphism (Section~\ref{sec:dm-sort-system}) corresponds to this construction but relaxes the injectivity requirement---an object can belong to multiple parent objects simultaneously, enabling cross-classification that strict subtypes do not permit.

Ologs, however, never produced a practical tool. Spivak acknowledged: ``writing a good olog demands a clarity of thought that ordinary writing or conversation can more easily elide''~\cite{Spivak2012}. Fifteen years later, the ``future directions'' outlined in Section~7.3 of the original paper remain unrealized---no interactive editor, no end-user application. A deeper limitation is that olog aspects are restricted to functional relations: each element of the source type maps to \emph{exactly one} element of the target type. Spivak explicitly marks non-functional relations as ``invalid aspects'' (\cite{Spivak2012}, Section~2.2.1). This excludes inherently one-to-many relationships central to knowledge management: a theorem may have multiple independent proofs, a concept may generalize in several directions, and a definition may be formalized by multiple complementary declarations.

Spivak's subsequent work on Functorial Data Migration~\cite{Spivak2012FDM} used functors to migrate data between database schemas. The core insight---that functors are the right abstraction for structure-preserving data integration---directly informs our plugin architecture, where import plugins are functors from external data source categories to the Astrolabe category.

More recent categorical approaches include multi-model database unification~\cite{Koupil2022} and the Topos Institute's work on double categories for knowledge representation~\cite{ToposInstitute}. Category-theoretic design patterns have also been applied in model-driven software engineering~\cite{DiskinMaibaum}. All remain at the theoretical level---none has produced an interactive knowledge management tool.

Astrolabe inherits ologs' core insight (objects and morphisms as first-class citizens) and its inter-olog communication architecture (functors as knowledge base connectors). We extend in three directions: (1)~the sort system replaces functional aspects with typed, many-to-many morphisms---what ologs mark as ``invalid'' becomes our primary modeling primitive; (2)~our \texttt{equivalent} morphism sort partially recovers the role of olog facts (commutative diagrams) by declaring path equivalence, with full commutative diagram support as future work; (3)~AI-assisted structuring, content-addressing, network analysis, and a plugin architecture add engineering dimensions that ologs did not address.


\subsection{Graph-Oriented Knowledge Management Tools}
\label{sec:bg-pkm}

A recent generation of personal knowledge management (PKM) tools---Roam Research~\cite{RoamResearch}, Obsidian~\cite{Obsidian}, Logseq~\cite{Logseq}, Tana, Capacities, and Heptabase---has brought graph-based organization into mainstream use. These systems vary in their core abstractions (block-level outliners, Markdown files, visual whiteboards), but all provide some form of bidirectional linking and graph visualization. We evaluate them along four dimensions relevant to our work: \emph{typed nodes}, \emph{typed edges}, \emph{edge identity} (whether edges are first-class entities with independent identifiers), and \emph{network analysis capability}.

Table~\ref{tab:pkm-comparison} summarizes the comparison. Several tools support typed \emph{nodes}---Logseq's property system, Tana's supertags~\cite{Tana}, and Capacities' object types all allow structured metadata on entities. However, none supports typed \emph{edges}: all relationships are undifferentiated links without semantic labels, directional meaning, or independent identity. No tool provides built-in network analysis (centrality, community detection, path analysis) over the link graph.

\begin{table}[h]
\centering
\caption{Graph capabilities of knowledge management tools. \checkmark\ = supported, $\circ$ = partial, \texttimes\ = absent.}
\label{tab:pkm-comparison}
\small
\begin{tabular}{lcccc}
\toprule
Tool & Typed nodes & Typed edges & Edge identity & Network analysis \\
\midrule
Obsidian        & \texttimes & \texttimes & \texttimes & \texttimes \\
Logseq          & $\circ$    & \texttimes & \texttimes & \texttimes \\
Roam Research   & \texttimes & \texttimes & \texttimes & \texttimes \\
Tana            & \checkmark & \texttimes & \texttimes & \texttimes \\
Capacities      & $\circ$    & \texttimes & \texttimes & \texttimes \\
Heptabase       & \texttimes & \texttimes & \texttimes & \texttimes \\
\midrule
Astrolabe       & \checkmark & \checkmark & \checkmark & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

The pattern is consistent: these tools have adopted graph \emph{visualization} but not graph \emph{semantics}. Their core abstraction remains the document---a page, block, or card---with links as by-products of cross-referencing. This architecture has a structural consequence: because all edges are identical in type and weight, standard network analysis algorithms produce uninformative results. Betweenness centrality on a graph where ``depends on,'' ``is associated with,'' and ``contradicts'' are all the same edge type conflates logically distinct dimensions.

The limitation is architectural rather than incremental. Retrofitting typed edges onto a document-centric system creates a dual data model---documents alongside a graph---that must be kept in sync, with every plugin required to understand the type system. In Astrolabe, this inversion is resolved at the foundation: obj/mor \emph{is} the primary data structure, and document content (MDX) is a field within an obj, not the container from which the graph is derived.
