\section{Discussion}
\label{sec:discussion}


\subsection{MDX as Content Format: Feasibility and Limitations}
\label{sec:disc-mdx}

MDX offers three advantages over \LaTeX{} as a content substrate for knowledge management: structural addressability (components reference graph entities by ID, not by label), interactivity (React components enable dynamic visualizations within the content), and AI compatibility (language models generate valid MDX more reliably than structured \LaTeX{}, particularly for diagrams).

The principal limitation is the academic publishing ecosystem. Journals and conferences require \LaTeX{} or PDF submissions; MDX has no established role in this pipeline. For Astrolabe to serve researchers who must ultimately produce \LaTeX{} papers, an MDX-to-\LaTeX{} export path is necessary. This is a tractable engineering problem---the obj/mor data model contains all the semantic information needed to generate \texttt{\textbackslash begin\{theorem\}} environments from \texttt{theorem}-sorted objects---but it is not yet implemented.

A deeper question is whether the source format for mathematical writing should be a typesetting language at all. \LaTeX{} conflates content structure with visual presentation; MDX separates them. As AI-assisted writing becomes more prevalent, the ability to programmatically consume and transform content---which MDX supports and PDF does not---may prove more valuable than fine-grained typographic control.


\subsection{Knowledge Structure Beyond Linear Documents}
\label{sec:disc-papers}

Academic papers serve three functions: organizing knowledge, attributing credit, and disseminating results. These functions are real, but the paper as a delivery format has a structural cost: it linearizes inherently networked knowledge into a one-dimensional sequence. Chapter ordering reflects the author's narrative choices, not the knowledge's intrinsic dependency structure. Two papers that prove the same theorem via different methods have no formal mechanism for expressing that their results are equivalent.

The obj/mor model represents knowledge structure directly. Objects are knowledge units with typed content; morphisms are typed relationships between them; the graph's topology encodes dependency, generalization, and equivalence without requiring a linear ordering. A ``paper'' in this framing is one possible linearization of a subgraph---a view, not the source of truth.

When objects carry content-addressed identifiers (Section~\ref{sec:dm-identifiers}), their identity becomes independent of any particular document, database, or institution. Two Astrolabe instances that independently create an object for the same theorem can merge via hash matching. This aligns with the decentralized formal mathematics vision developed in~\cite{LiParisSeverini2026}: content-addressing provides the identity layer, and the categorical data model provides the structural layer.


\subsection{Validity Boundaries of Network Analysis Signals}
\label{sec:disc-validity}

Network analysis on typed knowledge graphs produces signals---centrality rankings, community partitions, bottleneck scores---that can inform formalization strategy. These signals have clear validity boundaries.

Betweenness centrality identifies nodes that lie on many shortest paths, which in a dependency graph corresponds to results that mediate between many pairs of dependent declarations. High betweenness does not imply mathematical importance: a technical coercion lemma may have high centrality purely because it mediates type conversions across the library, without being mathematically interesting. Conversely, a deep theorem with few dependencies may have low centrality despite its significance.

The sort system mitigates this problem. Computing centrality on \texttt{uses} edges alone (statement-level dependencies) produces different rankings than computing on all edges including \texttt{proof-uses}. Sort-separated analysis can distinguish infrastructure nodes (high centrality on mixed edges) from mathematically central results (high centrality on \texttt{uses} edges specifically). The evaluation metrics in Section~\ref{sec:eval-metrics} are designed to measure this divergence.

Network analysis is an auxiliary signal for human decision-making, not an automatic prioritization algorithm. The value lies in surfacing structural patterns that are invisible when examining declarations individually---which nodes are bottlenecks, which clusters can be formalized independently, where the graph is dense versus sparse---rather than in producing a definitive ranking.


\subsection{Domain Generality}
\label{sec:disc-generality}

The obj/mor model with open sort vocabularies is not specific to mathematics. A legal knowledge graph might use sorts \texttt{statute}, \texttt{case}, \texttt{regulation} for objects and \texttt{interprets}, \texttt{overrules}, \texttt{cites} for morphisms. A biological pathway map might use \texttt{gene}, \texttt{protein}, \texttt{pathway} and \texttt{regulates}, \texttt{inhibits}, \texttt{encodes}. The sort system, automatic color assignment, network analysis plugins, and MDX content substrate all function identically regardless of the domain vocabulary.

This paper focuses on mathematical formalization because it provides the most demanding test case---formal verification creates a ground truth against which the data model's fidelity can be measured---and because the blueprint--autoformalization gap (Section~\ref{sec:intro-gap}) provides a concrete motivation. Domain generality is an architectural property of the model, not a claim we validate empirically in this work.


\subsection{Networked Annotation for AI Training}
\label{sec:disc-ai-annotation}

Current data preparation for AI-assisted formalization is point-wise: individual statement--proof pairs are labeled, single declarations receive tags, or isolated problems are collected into benchmarks. Each annotation is independent; it carries no information about how the annotated item relates to other items in the corpus.

Constructing an obj/mor network produces annotations at three levels simultaneously. Creating an object annotates a knowledge unit with its type (sort) and content fields. Creating a morphism annotates a \emph{relationship} between two objects---its sort label encodes the nature of the relationship (\texttt{uses}, \texttt{generalizes}, \texttt{formalizes}). Running analysis plugins annotates the object's \emph{structural position}---centrality scores, community membership, and depth in the dependency DAG---without any additional manual effort.

The result is that network analysis signals function as automatic annotations: they emerge from the graph structure that users create during normal knowledge management work, rather than requiring a separate labeling pass. A training pipeline that consumes Astrolabe's typed graph sees not just isolated statement--proof pairs but a dependency network with typed edges, weighted by centrality, partitioned into communities. Whether this richer structural context improves AI formalization performance is an empirical question we leave to future work.


\subsection{Limitations}
\label{sec:disc-limitations}

Several limitations constrain the current system. The plugin architecture (v1) provides no sandboxing; a malicious or buggy plugin has full access to the knowledge graph. The storage backend uses a single JSON file (\texttt{knowledge.json}), which limits scalability---Mathlib's 100{,}000+ declarations approach the practical ceiling for single-file I/O, and production use at that scale would require a sharded or database-backed storage layer. The \texttt{.ilean} format parsed by $F_{\mathrm{Lean}}$ is an internal compilation artifact of Lean~4, subject to change across Lean versions without stability guarantees.

On the modeling side, the current implementation does not support commutative diagrams (path equivalence declarations), which limits the expressiveness of the Astrolabe category relative to full ologs. The \texttt{equivalent} morphism sort provides a partial workaround for binary equivalences but does not capture the general case of $n$-path commutativity.


\subsection{Relationship to Prior Work by the Authors}
\label{sec:disc-prior-work}

This paper builds on two prior works by the authors. \emph{The Network Structure of Mathlib}~\cite{LiPengSeverini2026} developed the network science methodology and dataset used in our evaluation; Astrolabe integrates these analytical capabilities as plugins rather than standalone scripts. \emph{Toward Decentralized Formal Mathematics}~\cite{LiParisSeverini2026} established the content-addressing pipeline (canonical reduction, expression serialization, SHA-256 hashing) from which Astrolabe's two-layer identifier design derives. The independent contribution of the present paper is the categorical data model (Section~\ref{sec:data-model}), the closed-loop pipeline (Section~\ref{sec:dm-pipeline}), and the system implementation (Section~\ref{sec:system}).
