\section{The Astrolabe}
\label{sec:foundations}

We build the theory from a single starting point: a finite alphabet. Every externalized piece of knowledge is a finite string; structured knowledge is a finite sequence of key-value pairs of strings. From this minimal assumption we construct hash identifiers, a reference structure, and restriction maps, arriving at the \emph{Astrolabe} $(\Delta(\Sigma), h_{\mathrm{id}})$. Simplices in an Astrolabe are graded by the length of their reference list, yielding the graded set $\Delta(\Sigma) = \bigsqcup_{k \geq 0} \Delta_k$. The restriction maps satisfy $\partial^2 = 0$, giving a chain complex on which the homological algebra of \S\ref{sec:topology} operates unconditionally. The geometric structure (semantic embeddings, metrics) is introduced in \S\ref{sec:semantics}.

\subsection{Alphabet and String Space}

\begin{definition}[Alphabet]
\label{def:alphabet}
An \emph{alphabet} is a finite set $A$. We impose no structure on $A$ beyond finiteness; it may be the set of Unicode code points, ASCII characters, or any other finite encoding.
\end{definition}

\begin{definition}[Monoid]
\label{def:monoid}
A \emph{monoid} is a triple $(M, \cdot, e)$ where $M$ is a set, $\cdot : M \times M \to M$ is an associative binary operation, and $e \in M$ is an identity element: $e \cdot m = m \cdot e = m$ for all $m \in M$.
\end{definition}

\begin{definition}[String space]
\label{def:string-space}
The \emph{string space} $A^*$ is the free monoid on $A$: the set of all finite sequences of elements of $A$, with concatenation as the monoid operation and the empty sequence $\varepsilon$ as identity.
\end{definition}

Every \LaTeX\ file, every Lean source, every JSON document, and every natural-language paragraph is an element of $A^*$ for an appropriate choice of $A$ (e.g., $A_{\mathrm{ASCII}}$ with $|A| = 128$, or the full Unicode code point set).

\begin{theorem}[Universal property of $A^*$]
\label{thm:free-monoid}
For any monoid $(M, \cdot, e)$ and any function $f : A \to M$, there exists a unique monoid homomorphism $\bar{f} : A^* \to M$ extending $f$.
\end{theorem}

In Lean~4, $A^*$ is \texttt{FreeMonoid} in Mathlib, and the universal property is \texttt{FreeMonoid.lift}, a corollary of the adjunction $\mathrm{Free} \dashv U$ between the free monoid functor and the forgetful functor. The universal property guarantees that any structure-preserving interpretation of the alphabet extends uniquely to all strings; this is the algebraic foundation of parsing.

\begin{definition}[Record space]
\label{def:record-space}
A \emph{record} is a finite sequence of key-value pairs, where both keys and values are strings:
\[
  R = (A^* \times A^*)^*.
\]
A record $r = ((k_1, v_1), \ldots, (k_n, v_n)) \in R$ assigns to each key $k_i$ a value $v_i$. Keys may repeat; the order of pairs is significant.
\end{definition}

\subsection{Hash and Content Addressing}

\begin{definition}[Hash space]
\label{def:hash-space}
Let $A_{\mathrm{hex}} = \{0, \ldots, 9, a, \ldots, f\} \subset A$ be the hexadecimal alphabet. The \emph{hash space} is
\[
H = A_{\mathrm{hex}}^*,
\]
the set of finite hexadecimal strings. In particular, $H \subset A^*$.
\end{definition}

\begin{definition}[Reference space]
\label{def:reference-space}
The \emph{reference space} is
\[
H^+ = \bigsqcup_{m \geq 1} H^m,
\]
the set of all finite-length ordered tuples of hashes. An element $\texttt{ref} \in H^+$ with $|\texttt{ref}| = k+1$ is a \emph{vertex list} of length $k+1$.
\end{definition}

\begin{definition}[Content addressing]
\label{def:content-addressing}
A \emph{content-addressing function} is an injective function
\[
h_{\mathrm{id}} : H^+ \times R \to H
\]
that assigns to each pair $(\texttt{ref}, \texttt{record})$ a unique hash. The identity of a simplex is determined by \emph{what it connects} ($\texttt{ref}$) and \emph{what it says} ($\texttt{record}$), not by its location.
\end{definition}

This is the same principle underlying Git commits and IPFS~\cite{Benet2014}. The implementation of $h_{\mathrm{id}}$ (hash function, serialization, collision avoidance) is a system parameter discussed in \S\ref{sec:implementation}; the theory depends only on injectivity.

\begin{example}
In practice, $h_{\mathrm{id}}$ is computed by applying a cryptographic hash function (e.g.\ SHA-256) to the dimension-prefixed serialization of the vertex list and the record.

For a 0-simplex (vertex):
\begin{minted}[fontsize=\scriptsize]{python}
record = '{"name":"Nat.add_comm","sort":"theorem","stmt":"..."}'
h_id   = sha256("0:" + record).hexdigest()[:12]  # "a1b2c3"
ref    = [h_id]                                   # ["a1b2c3"]
\end{minted}
\noindent The 0-simplex's $\texttt{ref}$ contains its own hash, so the construction is self-referential: $h_{\mathrm{id}}^{(0)}$ is the unique fixed point of $h \mapsto \texttt{SHA256}(\texttt{"0:"} \| \texttt{record})$.

For a 1-simplex (edge) connecting two vertices:
\begin{minted}[fontsize=\scriptsize]{python}
ref    = '["a1b2c3","b2c3d4"]'
record = '{"sort":"uses","weight":"0.9"}'
h_id   = sha256("1:" + ref + record).hexdigest()[:12]  # "x7y8z9"
\end{minted}
\noindent The same two vertices with different $\texttt{record}$ produce a different hash; they are distinct 1-simplices. The dimension prefix ensures that a vertex hash and an edge hash never collide, even if their serialized content happens to match.
\end{example}

\begin{definition}[Parsing]
\label{def:parsing}
Let $\mathcal{P}_{\mathrm{fin}}(R)$ denote the set of finite subsets of $R$:
\[
\mathcal{P}_{\mathrm{fin}}(R) = \{ S \subseteq R \mid |S| < \infty \}.
\]
A \emph{parsing procedure} is a function $P : A^* \to \mathcal{P}_{\mathrm{fin}}(R)$ that maps a string to a finite set of records. Different parsing procedures extract different structures from the same string; we distinguish them by subscripts ($P_{\mathrm{compiler}}$, $P_{\mathrm{LLM}}$, etc.).
\end{definition}

\begin{example}
Consider the Lean source file $s \in A^*$:
\begin{minted}[fontsize=\scriptsize]{lean4}
theorem length_append (l1 l2 : List a) :
    (l1 ++ l2).length = l1.length + l2.length := by
  induction l1 <;> simp [*]
\end{minted}
\noindent Three parsers extract different sets of records from the same string:
\begin{itemize}
\item $P_{\mathrm{compiler}}(s)$: the Lean~4 compiler reads \texttt{.ilean} artifacts and produces one record per declaration. Writing each record as an element of $R = (A^* \times A^*)^*$:
\begin{align*}
r_1 &= ((\texttt{name},\; \texttt{length\_append}),\; (\texttt{sort},\; \texttt{theorem}),\; (\texttt{stmt},\; \ldots\,)) \in R, \\
r_2 &= ((\texttt{name},\; \texttt{List.length}),\; (\texttt{sort},\; \texttt{definition})) \in R, \\
r_3 &= ((\texttt{name},\; \texttt{List.append}),\; (\texttt{sort},\; \texttt{definition})) \in R, \\
r_4 &= ((\texttt{name},\; \texttt{Nat.add}),\; (\texttt{sort},\; \texttt{definition})) \in R.
\end{align*}
Thus $P_{\mathrm{compiler}}(s) = \{r_1, r_2, r_3, r_4\} \in \mathcal{P}_{\mathrm{fin}}(R)$.

\item $P_{\mathrm{LLM}}(s)$: a language model reads the same file as natural language and extracts records for each mathematical concept:
\[
P_{\mathrm{LLM}}(s) = \bigl\{\,
((\texttt{name},\; \texttt{list concatenation}),\; (\texttt{sort},\; \texttt{concept})),\;\ldots
\,\bigr\} \in \mathcal{P}_{\mathrm{fin}}(R).
\]

\item $P_{\mathrm{AST}}(s)$: a syntax-level parser extracts the abstract syntax tree, producing records for each node:
\[
P_{\mathrm{AST}}(s) = \bigl\{\,
((\texttt{name},\; \texttt{length\_append}),\; (\texttt{sort},\; \texttt{theorem})),\;\ldots
\,\bigr\} \in \mathcal{P}_{\mathrm{fin}}(R).
\]
\end{itemize}
\noindent In general $P_{\mathrm{compiler}}(s) \neq P_{\mathrm{LLM}}(s) \neq P_{\mathrm{AST}}(s)$: different parsers extract different records from the same string. The framework is indifferent to domain; the same construction applies to Python modules, legal statutes, or biological databases.
\end{example}

\subsection{Combinatorial Structure}

Let $\mathbb{A}$ be an \texttt{astrolabe.json} file. With $A$, $A^*$, $R$, $H$, $H^+$, and $h_{\mathrm{id}}$ in hand, we define the combinatorial objects that populate $\mathbb{A}$.

\begin{definition}[Semantic atom]
\label{def:atom}
A \emph{semantic atom} is an element $\sigma^0 \in H^+ \times H \times R$. Write $\texttt{ref}(\sigma^0)$, $\texttt{hash}(\sigma^0)$, $\texttt{record}(\sigma^0)$ for the three projections. A semantic atom satisfies:
\[
|\texttt{ref}(\sigma^0)| = 1, \qquad \texttt{ref}(\sigma^0) = [\texttt{hash}(\sigma^0)].
\]
\end{definition}

A semantic atom is the most basic entry in \texttt{astrolabe.json}. Its $\texttt{ref}$ contains only its own hash; it does not reference any other entry. All of its content lives in $\texttt{record}$.

\begin{example}
In \texttt{astrolabe.json}:
\begin{minted}[fontsize=\scriptsize]{json}
"a1b2c3": {
  "ref": ["a1b2c3"],
  "record": {"name": "Nat.add_comm", "sort": "theorem"}
}
\end{minted}
\end{example}

\begin{definition}[Semantic form]
\label{def:form}
A \emph{semantic form of degree $k$} ($k \geq 1$) is an element $\sigma^k \in H^+ \times H \times R$ where $\texttt{ref}(\sigma^k) \in H^{k+1}$.
\end{definition}

The case $k = 0$ is excluded: a single-element $\texttt{ref}$ forces $\texttt{ref} = [\texttt{hash}]$, which is the defining property of a semantic atom. Since $\texttt{ref} \in H^{k+1}$ is an ordered tuple, $[a, b] \neq [b, a]$; repetition is permitted, so $[a, a, b]$ is a legal $\texttt{ref}$.

\begin{lemma}[Pure/mixed decomposition]
\label{lem:pure-mixed}
Every semantic form is exactly one of:
\begin{itemize}
\item \emph{pure atom form}: every hash in $\texttt{ref}$ points to a semantic atom;
\item \emph{mixed form}: at least one hash in $\texttt{ref}$ points to another form.
\end{itemize}
\end{lemma}

\begin{proof}
By the law of excluded middle: either every hash in $\texttt{ref}$ points to an atom, or not.
\end{proof}

\begin{definition}[Stage]
\label{def:stage}
Given an \texttt{astrolabe.json} file $\mathbb{A}$, the \emph{stages} of $\mathbb{A}$ are defined inductively. Stage $0$ is the set of all semantic atoms in $\mathbb{A}$, with hash pool $P_0$:
\[
\mathcal{S}_0 = \bigl\{\sigma^0 \in H^+ \times H \times R \;\big|\; \texttt{ref}(\sigma^0) = [\texttt{hash}(\sigma^0)]\bigr\},
\qquad
P_0 = \bigl\{\texttt{hash}(\sigma) \;\big|\; \sigma \in \mathcal{S}_0\bigr\}.
\]
Stage $m+1$ adds all forms whose $\texttt{ref}$ hashes belong to the previous hash pool $P_m$:
\[
\mathcal{S}_{m+1} = \mathcal{S}_m \;\cup\; \bigl\{\sigma^k \in H^+ \times H \times R \;\big|\; k \geq 1,\; \texttt{ref}(\sigma^k) \in P_m^{k+1}\bigr\},
\qquad
P_{m+1} = \bigl\{\texttt{hash}(\sigma) \;\big|\; \sigma \in \mathcal{S}_{m+1}\bigr\}.
\]
\end{definition}

\begin{definition}[Content of \texttt{astrolabe.json}]
\label{def:astrolabe-content}
The set of all entries that may appear in \texttt{astrolabe.json} is $\bigcup_{m \geq 0} \mathcal{S}_m$.
\end{definition}

Each stage adds entries with a clear interpretation: $\mathcal{S}_0$ contains atomic knowledge, $\mathcal{S}_1$ adds direct relations between atoms, and $\mathcal{S}_{m+1}$ adds relations that reference entries up to $\mathcal{S}_m$.

\subsection{Semantic Simplex}

\begin{definition}[Signature]
\label{def:signature}
A \emph{signature} is a pair $\Sigma = (S, h_{\mathrm{id}})$ where $S$ is a finite set of simplices and $h_{\mathrm{id}} : H^+ \times R \hookrightarrow H$ is an injective content-addressing function. Since $h_{\mathrm{id}}$ is injective, it induces a dictionary structure: each simplex is uniquely indexed by its hash.
\end{definition}

\begin{definition}[Semantic simplex]
\label{def:semantic-simplex}
Let $\Sigma = (S, h_{\mathrm{id}})$ be a signature. A \emph{semantic simplex} is an element
\[
\sigma \in H^+ \times H \times R.
\]
If $|\texttt{ref}| = k+1$, we say $\sigma$ has dimension $k$ and write $\sigma_k$. The hashes in $\texttt{ref}$ may refer to simplices of any dimension; there is no constraint.
\end{definition}

Several features of this definition deserve emphasis.

\begin{enumerate}
\item \textbf{$\texttt{ref}$ is ordered.} The vertex list $[a, b]$ and $[b, a]$ are distinct simplices with distinct hashes. The ordering determines the signs in the restriction maps and hence in $\partial$. This is consistent with oriented simplices in classical topology.

\item \textbf{Dangling references are permitted.} A 1-simplex with $\texttt{ref} = [\texttt{abc}, \texttt{def}]$ is well-defined even if no 0-simplices with hashes $\texttt{abc}$ or $\texttt{def}$ exist in~$S$. The hash space $H$ does not track which hashes correspond to existing simplices. Whether to allow dangling references is an engineering choice, not a theoretical one.

\item \textbf{Repeated hashes are permitted.} The list $[a, a, b]$ defines a legal 2-simplex. Classical simplicial complexes forbid repeated vertices; here, the semantics is that $a$ participates twice in the relation. Note that $d_0$ and $d_1$ both produce $[a, b]$, so the corresponding terms in $\partial$ cancel.

\item \textbf{``Simplex'' is a borrowed name.} A classical $k$-simplex is the convex hull of $k+1$ affinely independent points in $\mathbb{R}^{k+1}$. A semantic simplex is not a convex hull, does not live in $\mathbb{R}^n$, and does not require affine independence. The name is borrowed because the restriction maps and $\partial^2 = 0$ share the same algebraic structure.

\item \textbf{Dimension is combinatorial, not geometric.} A 2-simplex with $\texttt{ref} = [h_1, h_2, h_3]$ has dimension~2 because $|\texttt{ref}| = 3$, regardless of whether $h_1, h_2, h_3$ point to 0-simplices, 1-simplices, or higher. The dimension of a simplex is determined by its own $\texttt{ref}$, not by the dimensions of its references.
\end{enumerate}

\begin{figure}[ht]
\centering
\begin{tabular}{|l|c|c|c|}
\hline
& $k = 0$ & $k = 1$ & $k = 2$ \\
\hline
% --- Row 1: geometry ---
\textbf{Simplex}
&
\begin{tikzpicture}[baseline=-5pt,
  v/.style={circle, fill, inner sep=0.8pt}]
\node[v] (p) at (0,0) {};
\node[font=\tiny, above=2pt] at (p) {\texttt{a1b2c3}};
\end{tikzpicture}
&
\begin{tikzpicture}[baseline=-5pt,
  vg/.style={circle, draw=gray!60, inner sep=0.8pt}]
\node[vg] (a) at (0,0) {};
\node[vg] (b) at (1.4,0) {};
\draw[semithick, blue!60!black] (a)--(b);
\node[font=\tiny, gray, above=2pt] at (a) {\texttt{a1b2c3}};
\node[font=\tiny, gray, above=2pt] at (b) {\texttt{b2c3d4}};
\end{tikzpicture}
&
\begin{tikzpicture}[baseline=-5pt,
  vg/.style={circle, draw=gray!60, inner sep=0.8pt}]
\coordinate (a) at (0,0);
\coordinate (b) at (1.4,0);
\coordinate (c) at (0.7,1.1);
\fill[red!15] (a)--(b)--(c)--cycle;
\node[vg] at (a) {};
\node[vg] at (b) {};
\node[vg] at (c) {};
\node[font=\tiny, gray, below left=1pt] at (a) {\texttt{a1b2c3}};
\node[font=\tiny, gray, below right=1pt] at (b) {\texttt{b2c3d4}};
\node[font=\tiny, gray, above=2pt] at (c) {\texttt{c3d4e5}};
\end{tikzpicture}
\\
\hline
% --- Row 2: h_id ---
$h_{\mathrm{id}}$
&
\texttt{\scriptsize a1b2c3}
&
\textcolor{blue!60!black}{\texttt{\scriptsize x7y8z9}}
&
\textcolor{red!40!black}{\texttt{\scriptsize q1r2s3}}
\\
\hline
% --- Row 3: JSON ---
\textbf{Data}
&
\begin{minipage}[t]{0.25\textwidth}
\begin{minted}[fontsize=\tiny,frame=none,bgcolor=gray!5]{json}
"a1b2c3": {
 "ref": ["a1b2c3"],
 "record": {
  "name": "Nat.add_comm",
  "sort": "theorem",
  "stmt": "..."
 }
}
\end{minted}
\end{minipage}
&
\begin{minipage}[t]{0.25\textwidth}
\begin{minted}[fontsize=\tiny,frame=none,bgcolor=gray!5]{json}
"x7y8z9": {
 "ref": ["a1b2c3",
         "b2c3d4"],
 "record": {
  "sort": "uses",
  "weight": "0.9"
 }
}
\end{minted}
\end{minipage}
&
\begin{minipage}[t]{0.25\textwidth}
\begin{minted}[fontsize=\tiny,frame=none,bgcolor=gray!5]{json}
"q1r2s3": {
 "ref": ["a1b2c3",
         "b2c3d4",
         "c3d4e5"],
 "record": {
  "sort": "joint_step",
  "method": "apply"
 }
}
\end{minted}
\end{minipage}
\\
\hline
\end{tabular}
\caption{Semantic simplices of dimension $0$, $1$, $2$. Gray circles denote referenced simplices, shown for context only. This example satisfies the classical constraint; in general, $\texttt{ref}$ may point to simplices of any dimension.}
\label{fig:simplices}
\end{figure}

Each simplex carries its own $h_{\mathrm{id}}$ and its own $\texttt{record}$. The 1-simplex \texttt{x7y8z9} records the \emph{relation} between two referenced simplices, not the content of those references; the 2-simplex \texttt{q1r2s3} records the \emph{ternary synergy}, not the content of its three references. The $\texttt{ref}$ field lists the hashes of the referenced simplices; the $\texttt{record}$ field is the simplex's own semantic content.

\begin{definition}[Restriction maps]
\label{def:restriction-maps}
For a $k$-simplex $\sigma$ with $\texttt{ref} = [v_0, \ldots, v_k]$, the \emph{$i$-th restriction map} is
\[
d_i(\sigma) = [v_0, \ldots, \hat{v}_i, \ldots, v_k], \qquad 0 \leq i \leq k,
\]
where $\hat{v}_i$ denotes omission of the $i$-th entry. Each $d_i$ is an information extractor: ``what remains if we forget participant $v_i$?'' The result is a vertex list of length $k$; the corresponding $(k-1)$-simplex may or may not exist in $S$.
\end{definition}

When $d_i(\sigma)$ does not exist in the knowledge base, a plugin may generate a new simplex with $\texttt{sort} = \texttt{"induced"}$. Subsequent plugins can enrich it; for instance, an LLM plugin may determine whether the sub-relation holds and update the sort to \texttt{"confirmed"} or \texttt{"independent"}.

\begin{proposition}[Restriction identity]
\label{prop:restriction-identity}
The restriction maps satisfy
\[
d_i \circ d_j = d_{j-1} \circ d_i \quad \text{for } i < j.
\]
\end{proposition}

\begin{proof}
Removing $v_j$ first and then $v_i$ ($i < j$): after removing $v_j$, the index of $v_i$ is unchanged (since $i < j$), so the second removal is $d_i$. Removing $v_i$ first and then $v_j$: after removing $v_i$, the index of $v_j$ shifts down by one to $j - 1$, so the second removal is $d_{j-1}$. Both operations yield the same list with $v_i$ and $v_j$ omitted.
\end{proof}

The restriction identity holds at the level of $\texttt{ref}$. It does not depend on what the hashes point to, what the records contain, or whether the referenced simplices exist. This is a purely combinatorial fact about ordered lists.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  v/.style={circle, fill=blue!70!black, inner sep=1.5pt},
  lbl/.style={font=\small, blue!70!black},
  arr/.style={->, >=Stealth, gray, thick}
]
\coordinate (a) at (0, 0);
\coordinate (b) at (2.2, 0);
\coordinate (c) at (1.1, 1.8);
\fill[blue!8] (a)--(b)--(c)--cycle;
\draw[semithick, blue!50!black] (a)--(b)--(c)--cycle;
\node[v] at (a) {};
\node[v] at (b) {};
\node[v] at (c) {};
\node[lbl, below left=1pt] at (a) {$v_0$};
\node[lbl, below right=1pt] at (b) {$v_1$};
\node[lbl, above=2pt] at (c) {$v_2$};
\node[font=\small] at (1.1, 0.5) {$\sigma$};
\node[font=\scriptsize, gray] at (1.1, -0.6) {$\texttt{ref} = [v_0, v_1, v_2]$};
\draw[arr] (3.0, 0.9) -- node[above, font=\small] {$d_0$} node[below, font=\scriptsize, red!60!black] {remove $v_0$} (5.0, 0.9);
\node[v] (r1) at (5.8, 0.4) {};
\node[v] (r2) at (7.3, 1.4) {};
\draw[semithick, blue!50!black] (r1)--(r2);
\node[lbl, below=2pt] at (r1) {$v_1$};
\node[lbl, above=2pt] at (r2) {$v_2$};
\node[font=\scriptsize, gray] at (6.9, -0.1) {$\texttt{ref} = [v_1, v_2]$};
\end{tikzpicture}
\caption{The restriction map $d_0$ applied to a 2-simplex $\sigma$: removing $v_0$ from $\texttt{ref} = [v_0, v_1, v_2]$ yields the vertex list $[v_1, v_2]$. Similarly, $d_1$ removes $v_1$ to give $[v_0, v_2]$, and $d_2$ removes $v_2$ to give $[v_0, v_1]$.}
\label{fig:restriction-maps}
\end{figure}

The restriction map $d_i$ answers the question: ``what remains if we forget the $i$-th vertex?'' For a 1-simplex $\sigma$ with $\texttt{ref} = [v_0, v_1]$:
\[
d_0(\sigma) = [v_1], \qquad d_1(\sigma) = [v_0].
\]
These are the two endpoints of the edge. In the free category (\S\ref{sec:category}), $d_1$ and $d_0$ recover the source and target maps: $s(\sigma) = v_0$, $t(\sigma) = v_1$.

For a 2-simplex $\sigma$ with $\texttt{ref} = [v_0, v_1, v_2]$:
\[
d_0(\sigma) = [v_1, v_2], \qquad d_1(\sigma) = [v_0, v_2], \qquad d_2(\sigma) = [v_0, v_1].
\]
These are the three edges of the triangle. The restriction $d_i$ is the edge \emph{opposite} to vertex $v_i$.

\begin{definition}[Boundary operator]
\label{def:boundary}
The \emph{boundary operator} $\partial_k : C_k \to C_{k-1}$ is defined on the free abelian group $C_k = \mathbb{Z}[\Delta_k]$ by
\[
\partial_k(\sigma) = \sum_{i=0}^{k} (-1)^i\, d_i(\sigma).
\]
\end{definition}

\begin{theorem}[$\partial^2 = 0$]
\label{thm:boundary-squared}
$\partial_{k-1} \circ \partial_k = 0$ for all $k$. This holds unconditionally: it does not depend on what $\texttt{ref}$ points to, what $\texttt{record}$ contains, or whether the referenced simplices exist.
\end{theorem}

\begin{proof}
By the restriction identity (Proposition~\ref{prop:restriction-identity}), each pair $(v_i, v_j)$ with $i < j$ appears twice in $\partial_{k-1} \circ \partial_k$ with opposite signs, hence cancels. Concretely:
\[
\partial_{k-1}(\partial_k(\sigma)) = \sum_{j=0}^{k} \sum_{i=0}^{k-1} (-1)^{i+j}\, d_i(d_j(\sigma)).
\]
Split into $i < j$ and $i \geq j$. For $i < j$, $d_i \circ d_j = d_{j-1} \circ d_i$; relabeling shows the two sums cancel term by term.
\end{proof}

This is the fundamental algebraic guarantee of the Astrolabe. The chain complex $\cdots \xrightarrow{\partial_{k+1}} C_k \xrightarrow{\partial_k} C_{k-1} \xrightarrow{\partial_{k-1}} \cdots$ is the starting point for all of \S\ref{sec:topology}.

\subsection{The Astrolabe}

\begin{definition}[Astrolabe]
\label{def:astrolabe}
An \emph{Astrolabe} is a pair $(\Delta(\Sigma), h_{\mathrm{id}})$ where $\Sigma = (S, h_{\mathrm{id}})$ is a signature and
\[
\Delta(\Sigma) = \bigsqcup_{k \geq 0} \Delta_k, \qquad \Delta_k = \{\sigma \in S \mid |\texttt{ref}(\sigma)| = k+1\}
\]
is the graded set of simplices. The set $\Delta_k$ is the \emph{$k$-stratum}. The \emph{$k$-skeleton} is $\mathrm{sk}_k(\Delta) = \bigsqcup_{j=0}^{k} \Delta_j$.
\end{definition}

The graded decomposition of $\Delta(\Sigma)$ induces a graded decomposition of the hash space:
\[
H^{(k)} = h_{\mathrm{id}}(\Delta_k) \subset H, \qquad H = \bigsqcup_{k \geq 0} H^{(k)}.
\]
The injectivity of $h_{\mathrm{id}}$ ensures that $H^{(j)} \cap H^{(k)} = \emptyset$ for $j \neq k$, since simplices in different strata have vertex lists of different lengths and thus different inputs to $h_{\mathrm{id}}$.

Not every position in the complex need be occupied by a substantive simplex. We explicitly record absences:

\begin{definition}[Semantic void]
\label{def:void}
A \emph{semantic void} is a simplex whose $\texttt{record}$ has a \texttt{sort} indicating absence rather than presence:
\begin{itemize}
\item $\texttt{sort} = \texttt{"independent"}$: confirmed absence of relation.
\item $\texttt{sort} = \texttt{"missing"}$: expected relation not yet recorded.
\item $\texttt{sort} = \texttt{"induced"}$: face induced by a higher simplex, semantics pending.
\item $\texttt{sort} = \texttt{"negated"}$: explicit negation of relation.
\end{itemize}
A void has its own $\texttt{hash}$ and is a full member of $\Delta(\Sigma)$. Absence is not a gap in the data structure but a record with semantic content.
\end{definition}

\subsection{Data Format: \texttt{astrolabe.json}}

The disk format is \texttt{astrolabe.json}, graded by dimension at the top level. Each entry is a triple $(\texttt{ref}, \texttt{hash}, \texttt{record})$.

\begin{table}[H]
\centering
\begin{tabular}{| C{0.3\textwidth} | C{0.3\textwidth} | C{0.3\textwidth} |}
\hline
\multicolumn{3}{|c|}{$\Delta(\Sigma)$} \\
\hline
\multicolumn{3}{|c|}{%
\begin{tikzpicture}[baseline=-5pt, scale=3, v/.style={circle, fill, inner sep=0.8pt}]
\node[v] (a) at (0.2,0.6) {};
\node[v] (b) at (1.8,0.1) {};
\node[v] (c) at (1.0,1.0) {};
\node[v] (d) at (2.5,0.8) {};
\fill[red!15] (a.center)--(b.center)--(c.center)--cycle;
\draw[semithick, blue!60!black] (c)--(b);
\draw[semithick, blue!60!black] (b)--(d);
\node[v] at (a) {};
\node[v] at (b) {};
\node[v] at (c) {};
\node[v] at (d) {};
\node[font=\tiny, above=2pt] at (a) {\texttt{a1b2c3}};
\node[font=\tiny, above=2pt] at (b) {\texttt{b2c3d4}};
\node[font=\tiny, above=2pt] at (c) {\texttt{c3d4e5}};
\node[font=\tiny, above=2pt] at (d) {\texttt{d4e5f6}};
\end{tikzpicture}%
} \\
\hline
\centering $\Delta_0$ & \centering $\Delta_1$ & \centering $\Delta_2$ \tabularnewline
\hline
\begin{tikzpicture}[baseline=-5pt, v/.style={circle, fill, inner sep=0.8pt}]
\node[v] (a) at (0.2,0.6) {};
\node[v] (b) at (1.8,0.1) {};
\node[v] (c) at (1.0,1.0) {};
\node[v] (d) at (2.5,0.8) {};
\node[font=\tiny, above=2pt] at (a) {\texttt{a1b2c3}};
\node[font=\tiny, above=2pt] at (b) {\texttt{b2c3d4}};
\node[font=\tiny, above=2pt] at (c) {\texttt{c3d4e5}};
\node[font=\tiny, above=2pt] at (d) {\texttt{d4e5f6}};
\end{tikzpicture}
 &
\begin{tikzpicture}[baseline=-5pt, vg/.style={circle, draw=gray!60, inner sep=0.8pt}]
\coordinate (a) at (0.2,0.6);
\node[vg] (b) at (1.8,0.1) {};
\node[vg] (c) at (1.0,1.0) {};
\node[vg] (d) at (2.5,0.8) {};
\draw[semithick, blue!60!black] (c)--(b);
\draw[semithick, blue!60!black] (b)--(d);
\node[font=\tiny, gray, above=2pt, opacity=0] at (a) {\texttt{a1b2c3}};
\node[font=\tiny, gray, above=2pt] at (b) {\texttt{b2c3d4}};
\node[font=\tiny, gray, above=2pt] at (c) {\texttt{c3d4e5}};
\node[font=\tiny, gray, above=2pt] at (d) {\texttt{d4e5f6}};
\end{tikzpicture}
 &
\begin{tikzpicture}[baseline=-5pt, vg/.style={circle, draw=gray!60, inner sep=0.8pt}]
\node[vg] (a) at (0.2,0.6) {};
\node[vg] (b) at (1.8,0.1) {};
\node[vg] (c) at (1.0,1.0) {};
\coordinate (d) at (2.5,0.8);
\fill[red!15] (a.center)--(b.center)--(c.center)--cycle;
\node[vg] at (a) {};
\node[vg] at (b) {};
\node[vg] at (c) {};
\node[font=\tiny, gray, above=2pt] at (a) {\texttt{a1b2c3}};
\node[font=\tiny, gray, above=2pt] at (b) {\texttt{b2c3d4}};
\node[font=\tiny, gray, above=2pt] at (c) {\texttt{c3d4e5}};
\node[font=\tiny, gray, above=2pt, opacity=0] at (d) {\texttt{d4e5f6}};
\end{tikzpicture}
 \\
\hline
\end{tabular}
\end{table}

Since $\texttt{ref}$ may reference hashes of any dimension, an Astrolabe can itself be recorded as a simplex: its $\texttt{ref}$ lists the hashes of all its members, and its $\texttt{record}$ carries metadata such as name and version. Complexes can reference other complexes, enabling hierarchical composition.

\begin{example}[Complex as simplex]
A small complex and its self-referencing entry in \texttt{astrolabe.json}:
\begin{verbatim}
{
  "0": {
    "a1b2c3": {
      "ref": ["a1b2c3"],
      "record": {"name": "Nat.add_comm", "sort": "theorem"}
    },
    "b2c3d4": {
      "ref": ["b2c3d4"],
      "record": {"name": "Nat.zero_add", "sort": "theorem"}
    }
  },
  "1": {
    "e1f2a3": {
      "ref": ["a1b2c3", "b2c3d4"],
      "record": {"sort": "uses"}
    }
  },
  "complex": {
    "cx1": {
      "ref": ["a1b2c3", "b2c3d4", "e1f2a3"],
      "record": {
        "sort": "complex",
        "name": "Nat_basics",
        "version": "1.0"
      }
    }
  }
}
\end{verbatim}
The entry \texttt{cx1} references all members of the complex (two vertices and one edge). Its hash is computed by $h_{\mathrm{id}}$ from these three member hashes together with its record. Modifying any member changes \texttt{cx1}'s hash, providing automatic version tracking.

Complexes can reference other complexes:
\begin{verbatim}
"complex": {
  "cx_all": {
    "ref": ["cx1", "cx2", "cx3"],
    "record": {
      "sort": "complex",
      "name": "Mathlib",
      "version": "4.3"
    }
  }
}
\end{verbatim}
This is the recursive application of content addressing: not only simplices but also complexes are identified by their content. Object knowledge and meta-knowledge coexist in the same \texttt{astrolabe.json} and share the same algebraic-topological machinery.
\end{example}

\subsection{Constraints and Classical Structure}

\begin{definition}[Classical constraint]
\label{def:classical-constraint}
An Astrolabe $\Delta(\Sigma)$ satisfies the \emph{classical constraint} if
\[
\texttt{ref}(\sigma) \in H^{(0)^{k+1}} \quad \text{for all } \sigma_k \in \Delta(\Sigma),
\]
i.e., every hash in $\texttt{ref}$ belongs to a 0-simplex. A 0-simplex has $\texttt{ref} = [\texttt{hash}]$: it references only itself and is an atom. The classical constraint says: you reference atoms only.
\end{definition}

Under the classical constraint, the restriction maps $d_i$ reduce to the classical \emph{face maps} of simplicial topology, and $\Delta(\Sigma)$ becomes a semi-simplicial set.

\begin{table}[H]
\centering
\begin{tabular}{ll}
\hline
\textbf{Constraint} & \textbf{Structure obtained} \\
\hline
No constraint & Astrolabe (most general) \\
$\texttt{ref} \subseteq H^{(0)}$ & Semi-simplicial set \\
$\texttt{ref} \subseteq H^{(0)}$ + all faces exist & Simplicial complex \\
$\texttt{ref} \subseteq H^{(0)}$ + $|\texttt{ref}| \leq 2$ & Directed graph \\
\hline
\end{tabular}
\end{table}

The three main sections of this paper correspond to three layers of tools, each requiring different constraints:

\begin{table}[H]
\centering
\small
\begin{tabular}{lccc}
\hline
\textbf{Tool} & \textbf{Classical?} & \textbf{$h_{\mathrm{emb}}$?} & \textbf{Section} \\
\hline
$\partial^2 = 0$ & No & No & \S\ref{sec:foundations} \\
$H_k$ definable and computable & No & No & \S\ref{sec:topology} \\
Long exact sequence & No & No & \S\ref{sec:topology} \\
Snake lemma & No & No & \S\ref{sec:topology} \\
Category $\mathbf{Ast}$ & No & No & \S\ref{sec:category} \\
Universal property (plugins) & No & No & \S\ref{sec:category} \\
Self-enrichment & No & No & \S\ref{sec:foundations}, \S\ref{sec:implementation} \\
\hline
$H_k$ geometric interpretation & \textbf{Yes} & No & \S\ref{sec:topology} \\
\hline
Exterior algebra embedding & \textbf{Yes} & \textbf{Yes} & \S\ref{sec:semantics} \\
Embedding obstruction & \textbf{Yes} & \textbf{Yes} & \S\ref{sec:semantics} \\
Hallucination metric $|\mathrm{Hal}|$ & \textbf{Yes} & \textbf{Yes} & \S\ref{sec:semantics} \\
Weighted homology & \textbf{Yes} & \textbf{Yes} & \S\ref{sec:semantics} \\
\hline
\end{tabular}
\end{table}

In the implementation, \texttt{astrolabe.json} stores the full Astrolabe (no constraint). Code extracts the sub-structure satisfying the classical constraint and runs geometric tools on it. Non-classical simplices remain in the file; categorical tools and self-enrichment do not require constraints. One file, one line of code to switch perspectives.

\subsection{Self-Enrichment}

Every computation result can be written back as a new simplex, enlarging the Astrolabe:

\begin{table}[H]
\centering
\small
\begin{tabular}{llll}
\hline
\textbf{Operation} & \textbf{Input} & \textbf{Output sort} & \textbf{Classical?} \\
\hline
Restriction $d_i$ & $k$-simplex & \texttt{induced\_face} & Yes \\
LLM embedding & 0-simplex & \texttt{embedding} & Yes \\
Wedge embedding & $k$-simplex & \texttt{wedge\_embedding} & Yes \\
PageRank & 0-simplex & \texttt{pagerank} & Yes \\
Hallucination risk & pair $(a,b)$ & \texttt{hal\_risk} & Yes \\
Homology class & cycle & \texttt{homology\_class} & Yes \\
Snake $\delta$ & $\ker h$ element & \texttt{connecting} & Yes \\
\hline
Complex record & all members & \texttt{complex} & No \\
Edge-face relation & edge + face & \texttt{meta} & No \\
\hline
\end{tabular}
\end{table}

The first seven operations preserve the classical constraint; the last two produce simplices whose $\texttt{ref}$ mixes dimensions.

The enrichment process is a monotonically increasing sequence:
\[
\Delta(\Sigma_0) \xrightarrow{P_1} \Delta(\Sigma_1) \xrightarrow{P_2} \Delta(\Sigma_2) \xrightarrow{P_3} \cdots
\]
where each $P_i$ is a plugin that reads the current Astrolabe and writes new simplices. The sequence $\Delta(\Sigma_0) \subseteq \Delta(\Sigma_1) \subseteq \cdots$ is a filtration, and its persistent homology (\S\ref{sec:topology}) tracks how the knowledge structure evolves under enrichment.
