% !TEX root = main.tex
% !TEX options = -shell-escape
\documentclass[10pt]{article}
\usepackage{preamble}

\title{Homological Algebra and Embedding Theory\\in the Structured Knowledge Management Problem\\[6pt]\large DRAFT}
\author{
  Xinze Li\thanks{Department of Mathematics, University of Toronto}
}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Every piece of structured knowledge is a finite string over a finite alphabet. We define the \emph{astrolabe file}: a finite subset $\mathbb{A} \subset H^+ \times H \times R$ of semantic simplices whose reference fields are unconstrained. The \emph{multiplicity profile} $\boldsymbol{\mu}(\sigma)$ records how many times each vertex appears in a reference list, decomposing each stage layer into its finest combinatorial partition. Grouping restriction maps by vertex yields $N$ boundary operators $\partial^{(1)}, \ldots, \partial^{(N)}$ that pairwise anticommute (\emph{multicomplex theorem}), making each stage layer a $\mathbb{Z}^N$-graded multicomplex. The classical identity $\partial^2 = 0$ is a single linear combination of the $\binom{N+1}{2}$ anticommutativity equations; homology, long exact sequences, and the snake lemma follow as corollaries.

Adding the classical constraint (all references point to vertices) recovers the geometric interpretations of algebraic topology. Adding LLM embeddings gives a \emph{stage embedding tower} $W_{m+1} = \bigwedge W_m$ with exterior algebra geometry; topology and geometry coexist, and their inconsistency pinpoints structural gaps. We prove an \emph{embedding obstruction theorem}: the topological complexity of a knowledge base gives a lower bound on embedding dimension, and give formal characterizations of hallucination, scaling saturation, ``understanding vs.\ memorization,'' and chain-of-thought reasoning.

Merging knowledge bases is not a special operation: disjoint union followed by cross-references, with homological invariants tracking the structural effect. The algebraic core is mechanically verified in Lean~4 (zero \texttt{sorry}). The data format \texttt{astrolabe.json} and a reference implementation are open-source.
\end{abstract}

\tableofcontents
\newpage 
\input{sections/introduction}
\input{sections/foundations}
\input{sections/topology}
\input{sections/semantics}
\input{sections/implementation}
\input{sections/conclusion}

\bibliographystyle{alpha}
\bibliography{references}

\newpage
\appendix
\input{sections/appendix}

\end{document}